---
title: "Alonso_Week 2 Homework Assignment"
author: "Martin Alonso"
date: "7/17/2018"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(dplyr)
require(scales)
```

IST687 Introduction to Data Science: Week 2 Homework.

Explore	the	mtcars	dataset	(which	is	already	included	in	R). Copy	the	mtcars	dataset	into	a
new	variable	(called	it	myCars),	so	that	if	you	mess	up,	you	can	start	again	very	easily	(by	copying	mtcars	into	myCars	again).

```{r cars}
cars <- rownames(mtcars)
myCars <- cbind(cars, mtcars)
```

# Step 1: What is the hp (hp stands for "horse power")

1. What is the highest hp?

```{r myCars}
max(myCars$hp)
```

2. Which car has the highest hp?

```{r}
myCars %>% filter(hp == max(hp))
```

# Step 2: Explore mgp (mpg stands for "miles per gallon")

3. What is the highest mpg?

```{r}
max(myCars$mpg)
```

4. Which car has the highest mpg?

```{r}
myCars %>% filter(mpg == max(mpg))
```

5. Create a sorted dataframe, based on mpg

```{r}
myCars %>% arrange(mpg) %>% head()
```

# Step 3. Which car has the "best" combination of mpg and hp?

6. I rescaled mpg and hp to 100, and obtained the average of the two new columns.
```{r}
myCarsBest <- myCars %>% mutate(
  mpg_rescale = rescale(mpg, to = c(0,100)),
  hp_rescale = rescale(hp, to = c(0,100)), 
  rank = (mpg_rescale + hp_rescale)/2
)

myCarsBest %>% head(n=3)

myCarsBest %>% filter(rank == max(rank))
```

7. The "best" car using this logic is the Maserati Bora.

# Step 4. Which car has the "best" combination of mpg and hp, where mpg and hp must be given equal weight?

```{r}
myCarsEqual <- myCars %>% 
  mutate(weighted_mean = mpg/mean(mpg) * hp/mean(hp)) %>% 
  ungroup()

myCarsEqual %>% head(n=3)

myCarsEqual %>% select(cars, mpg, hp, weighted_mean) %>% filter(weighted_mean == max(weighted_mean))
```

Giving similar weights to both mpg and hp, the Maserati Bora is still the best car available.

# Sources
1. Stack Exchange, https://stats.stackexchange.com/questions/25894/changing-the-scale-of-a-variable-to-0-100, 2018-07-17