---
title: "Alonso_Week 2 Lab Assignment"
author: "Martin Alonso"
date: "7/17/2018"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(dplyr)
```

IST687 Introduction to Data Science: Week 2 Lab.

The	“dataframe”	is	one	of	the	most	essential	data	structures	used	in	R.	It	is conceptually	equivalent	to	a	database	“relation”	and	to	the	typical	rectangular	dataset	with	variables	as columns	and	cases	as	rows.	For	this	activity,	you	will	gain	some	skill	with	manipulating	a dataframe.

# Task 1

Using the mtcars dataset, create a new column that measures engine displacement per cylinder in cubic inches. 

```{r cars}
mt_mycars <- mtcars

mt_mycars$disp_per_cyl <- mt_mycars$disp/mt_mycars$cyl

summary(mt_mycars$disp_per_cyl)
```

# Task 2

Create four vectors of friends or family members which include the person ID, the number of Pets, the order of their birth, and the number of siblings.
Group the vectors into a data frame called myFriends.

```{r}
personID <- c('Wife','Dad','Mum','Daughter','Mother-in-law')
Pets <- c(1, 7, 0, 0, 1)
Order <- c(1, 3, 1, 1, 2)
Siblings <- c(0, 2, 3, 0, 1)

myFriends <- data.frame(personID, Pets, Order, Siblings)
```

Show the structure of myFriends...

```{r}
str(myFriends)
```

...and the summary.

```{r}
summary(myFriends)
```

Finally, display the values using the `$` sign.
```{r}
myFriends$personID
myFriends$Pets
myFriends$Order
myFriends$Siblings
```
