---
title: "Alonso_Week 1 Lab Assignment"
author: "Martin Alonso"
date: "7/9/2018"
output: 
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

IST687 Introduction to Data Science: Week 1 Lab.

Boolean logic is logic where the answer to a question asked can only be `TRUE` or `FALSE`. There are no other options. 
Within R, there are three operators that are commonly used for Boolean logic: AND (`&`), OR (`|`), and NOT (`!`). 

# AND (`&`)
The element-wise AND (`&`) operator is used to test whether two elements (including all elements within a vector) are equal to each other. 

```{r}
# 1 & 1 is TRUE because they share the same element.
1 & 1 

# TRUE & TRUE are also TRUE because they share the same element
TRUE & TRUE

# TRUE & FALSE or FALSE & FALSE do not share the same element, hence they are both FALSE
TRUE & FALSE
FALSE & FALSE
```

In R, there also exist another Logical AND (`&&`) that only tests whether the first element of two vectors are both TRUE.
```{r}
# Returns TRUE since both of the first elements of the vectors are TRUE..
a <- c(1, 2, 3, 4)
b <- c(5, 6, 7, 8)
print(a && b)

# Returns FALSE because both of the first elements of the two vectors isn't TRUE
a <- c(1, 2, 3, 4)
d <- c(0, 6, 7, 8)
print(a && d)
```

# OR (`|`)
The element-wise OR (`|`) operator tests whether an element (or all elements in a vector) is either `TRUE` or `FALSE` based on whether one of them is `TRUE`.
```{r}
# Two integers are both TRUE values, hence the operator recognizes them as TRUE.
3 | 4

# The same can be said of 1 and TRUE.
1 | TRUE

# But if none of the options is TRUE, then the logical operator gives FALSE
0 | 0
FALSE | FALSE

```

There is also a Logical OR (`||`) operator that tests whether the first element of two vectors is TRUE or FALSE.
```{r}
# Returns TRUE since one of the first elements of the vectors are TRUE.
a <- c(1, 2, 3, 4)
b <- c(5, 6, 7, 8)
print(a || b)

# Returns FALSE because one of the first elements of the two vectors isn't TRUE
a <- c(FALSE, 2, 3, 4)
d <- c(0, 6, 7, 8)
```

# NOT (`!`)
The final operator is the NOT (`!`) operator that takes each element of a vector and negates it, giving it its opposite value.
```{r}
# All numbers, except 0, are TRUE. By negating a, we turn it FALSE.
a <- 1
!a 

# In the case of TRUE and FALSE, negating them inverses their values.
b <- c(TRUE, FALSE)
!b 
```

